\begin{abstract}
The dynamic fluctuation of backhaul capacity represents a critical performance bottleneck in 6G cell-free massive multiple-input multiple-output (MIMO) systems. To address this challenge, this paper proposes IB-CellFree-GNN, a novel task-oriented successive refinement framework rooted in the Variational Information Bottleneck principle. The framework introduces two core innovations to optimize resource allocation and signal detection. First, a lightweight, Gumbel-Softmax-driven policy network is deployed at distributed access points (APs) to dynamically infer the optimal quantization bit-width (0, 2, or 4 bits) based on local signal-to-noise ratios, effectively discarding irrelevant signals to preserve backhaul resources. Second, a bit-aware heterogeneous graph neural network (GNN) is designed at the central processing unit to perform robust spatial aggregation and multi-user interference cancellation, adaptively compensating for asymmetric quantization noise. Extensive simulations demonstrate that the proposed method achieves near-optimal 32-bit full-precision detection performance while consuming an average of only 2.5 bits per payload. Under heavy user load conditions, IB-CellFree-GNN reduces the bit error rate by 67.3\% compared to conventional fixed 2-bit quantization schemes. Furthermore, the system maintains highly reliable communication even under an extreme 50\% link failure rate. Finally, the artificial intelligence decision module exhibits an O(1) computational complexity scaling with respect to the number of AP antennas, confirming its exceptional scalability and robustness for practical 6G deployments.
\end{abstract}