\begin{abstract}
In 6G cell-free massive multiple-input multiple-output (MIMO) systems, the dynamic fluctuation of fronthaul and backhaul capacity imposes a fundamental performance bottleneck. Conventional fixed-rate quantization schemes lack the flexibility to adapt to instantaneous channel conditions and specific task objectives, resulting in significant performance degradation under heavy interference or link failures. To address these challenges, this paper proposes a resource-inference graph neural network (GNN) framework that incorporates a task-oriented successive refinement approach. Specifically, a Gumbel-Softmax-based bit-width policy network is deployed at the access points (APs) to enable adaptive discrete quantization (e.g., 0, 2, and 4 bits). This is jointly optimized with a heterogeneous GNN at the central processing unit, which performs quantization-aware detection and spatial interference cancellation. By dynamically allocating bit-widths according to local channel conditions, the framework preserves critical demodulation information while minimizing backhaul overhead. Numerical evaluations demonstrate that the proposed method achieves near-full-precision performance with an average of only 2.5 bits per AP. In heavy-load scenarios, the proposed approach reduces the bit error rate by 67.3\% compared to conventional fixed 2-bit quantization schemes. Furthermore, the system exhibits substantial robustness, maintaining a bit error rate of less than 2.2\% even under a 50\% AP dropout rate. Finally, the policy network demonstrates highly scalable, antenna-independent computational complexity, maintaining a constant computational overhead regardless of the antenna array size, thereby facilitating practical large-scale deployments.
\end{abstract}