Cell-free massive multiple-input multiple-output (MIMO) is a promising architecture for sixth-generation (6G) networks; however, its practical deployment is constrained by the limited and fluctuating bandwidth of fronthaul and midhaul links. This paper introduces IB-CellFree-GNN, a task-oriented successive refinement framework designed to optimize signal detection under stringent backhaul constraints. The proposed approach integrates a variational information bottleneck (VIB)-based progressive quantizer with a graph neural network (GNN) to facilitate joint quantization and detection. By employing a Gumbel-Softmax-based resource-inference module, the system enables differentiable bit-width allocation, allowing access points (APs) to adaptively prioritize critical information based on instantaneous channel conditions and global bandwidth limits. Numerical evaluations demonstrate that IB-CellFree-GNN achieves performance comparable to full-precision systems with an average of 2.5 bits per sample. In high-interference scenarios involving 12 users and 16 APs, the framework yields a 67.3\% reduction in bit error rate (BER) relative to conventional fixed 2-bit quantization. Furthermore, the system demonstrates robustness by maintaining a BER below 5\% even under extreme conditions where 50\% of the APs experience complete backhaul failure. These results highlight the efficacy of the task-oriented approach in enhancing resource efficiency and reliability for distributed 6G communication systems.